Robust Bayesian Analysis of Causal Inference Problems

### 2022-10-27 JASA-template.tex
some update

In causal inference, we are interested in estimating the causal
effect of independent variables on a dependent variable. Ideally,
randomised trials are the most efficient way to perform this task.
However, this is not very practical for several reasons; ethical 
concerns, design cost, population size, to name a few. This
leaves us with observational studies which are usually obtained
by means collecting data though surveys or record keeping.

Let an observational study give us the outcomes $Y=(Y_1$, \dots, $Y_n)$ along with 
corresponding treatment indicators $T=(T_1$, \dots, $T_n)$. Then the
individual causal
effect of the treatment $T_i$ on outcome $Y_i$ is given by:
\begin{align}
\delta \coloneqq E(Y\mid T=1) - E(Y\mid T=0).
\end{align}
However, we can not observe $(Y_i\mid T_i=1)$ and $(Y_i\mid T_i=0)$
simultaneously for the $i$-th individual. Instead, we can estimate
average causal effect of the treatment $T$ so that
\begin{align}
\hat{\delta} \coloneqq 
\frac{\sum_{i=1}^n(Y_i\mid T_i=1) - (Y_i\mid T_i=0)}{n}.
\end{align}
Now, let $X\coloneqq$ $[X_1$, \dots, $X_n]^T$ 
denote the $p$-dimensional covariates present in the studies and $\beta \coloneqq (\beta_1$, \dots, $\beta_p)$ denotes the vector of regression
coefficients. Then we can define a linear model for the outcome
so that
\begin{equation}
    Y_i = \beta_{T} T_i + \beta_0 + X\beta + \epsilon_i
\end{equation}
where $\epsilon_i\sim \mathcal{N}(0, \sigma^2)$. Clearly, when
the underlying true outcome model is linear,
\begin{equation}
    \delta = \beta_{T}\quad\text{and hence } 
    \hat{\delta} - \delta = \hat{\beta}_T - \beta_T.
\end{equation}

\begin{figure}
    \centering
    \begin{tikzpicture}[roundnode/.style={circle, draw=green!60, fill=green!5, very thick, minimum size=7mm},
squarednode/.style={rectangle, draw=red!60, fill=red!5, very thick, minimum size=5mm}]
    \node (1) at (0,0) {Treatment ($T$)};
    \node (2) [right = of 1] {Outcome ($Y$)};
    \node (4) [below = of 1] {Input ($X$)};

    \path (1) edge[->]  (2);
    \path (1) edge[<-]  (4);
    \path (2) edge[<-] (4);
\end{tikzpicture}
\caption{Confounding in causal models.}
    \label{fig:my_label}
\end{figure}

@@

\section{Robust Bayesian Analysis}
\subsection{Hierarchical model}
We perform a robust Bayesian analysis to obtain robust estimates. To do so, we propose the
following spike and slab priors \cite{ishwaran2005} to specify $\beta$ and $\gamma$.
For $1\le j\le p$,
\begin{align}
    \beta_j \mid \pi_{\beta}, \sigma^2 &\sim 
    \pi_{\beta}\normal\left(0, \tau_1^2\sigma^2\right) 
    + (1-\pi_{\beta}) \normal\left(0, \tau_0^2\sigma^2\right)\\
    \gamma_j \mid \pi_{\gamma} &\sim 
    \pi_{\gamma}\normal\left(0, \tau_1^2\right) 
    + (1-\pi_{\gamma}) \normal\left(0, \tau_0^2\right)\\
    \alpha\mid \sigma^2 &\sim \normal(0, \sigma^2)\\
    \sigma^2&\sim \text{InvGamma}(a, b)\\
    \pi_{\beta}, \pi_{\gamma}&\sim\text{Beta}\left(sq, s(1-q)\right).
\end{align}
We fix sufficiently small $\tau^2_0$
$(1\gg\tau_0^2>0)$ so that  $(\beta_j, \gamma_j) = (0,0)$ has its probability mass 
concentrated around zero. Therefore, this represents the spike component of our prior specification. 
To construct the slab component, we consider $\tau_1^2$ to be large so that 
$\tau_1^2\gg\tau_0^2$. This allows the prior for $(\beta_j, \gamma_j)\not=(0,0)$ to be flat.

>>>>

In causal inference, we are interested in estimating the causal
effect of independent variables on a dependent variable. Ideally,
randomised trials are the most efficient way to perform this task.
However, this is not very practical for several reasons; ethical 
concerns, design cost, population size, to name a few. This
leaves us with observational studies which are usually obtained
by means collecting data though surveys or record keeping. But this
can be problematic in the presence of confounders. That is when
the variables are associated with both the treatment and the outcome.
In such cases, we need to be extra cautious as otherwise it will
lead to unwanted bias in the treatment effect estimator \cite{rosenbaum83}.

Several works have been done in order to tackle the presence of
confounder variables. One such work in the topic was 
by \citet{Robins1986ANA} where the author used  a graphical
approach for the identification of the causal parameters.
\citet{rosenbaum1985} suggested the use of a link model to estimate
the propensity scores for all individuals. Later on several other
methods have been proposed based on propensity score matching.
A brief review on such methods can be found in \cite{winship99, stuart10}.

Bayesian approach in causal effect estimation has become a popular
topic in recent days. However, one of the earlier works on this can be
found in \cite{rubin1978}. Lately with the notion of variable selection
of high dimensional problems, Bayesian methodologies have become
more significant.

\begin{figure}
	\centering
	\begin{tikzpicture}[params/.style={circle, draw=green!60, fill=green!5, very thick, minimum size=7mm}]
	\node (1) at (0,0) {Treatment ($T$)};
	\node (2) [right = of 1] {Outcome ($Y$)};
	\node (4) [below = of 1] {Input ($X$)};

	\path (1) edge[->]  (2);
	\path (1) edge[<-]  (4);
	\path (2) edge[<-] (4);
	\end{tikzpicture}
	\caption{Confounding in causal models.}
	\label{fig:confounding}
\end{figure}

Let an observational study give us the outcomes $Y=(Y_1$, \dots, $Y_n)$ along with 
corresponding treatment indicators $T=(T_1$, \dots, $T_n)$. Then the
individual causal
effect of the treatment $T_i$ on outcome $Y_i$ is given by:
\begin{align}
\delta \coloneqq E(Y\mid T=1) - E(Y\mid T=0).
\end{align}
However, we can not observe $(Y_i\mid T_i=1)$ and $(Y_i\mid T_i=0)$
simultaneously for the $i$-th individual. Instead, we can estimate
average causal effect of the treatment $T$ so that
\begin{align}
\hat{\delta} \coloneqq 
\frac{\sum_{i=1}^n(Y_i\mid T_i=1) - (Y_i\mid T_i=0)}{n}.
\end{align}
This relies on an important assumption that the treatment effect
on the $i$-th subject given that they received the treatment is
equal to the treatment effect when they remain as the control
\cite{winship99}.

Now, let $X\coloneqq$ $[X_1$, \dots, $X_n]^T$ 
denote the $p$-dimensional covariates present in the studies and $\beta \coloneqq (\beta_1$, \dots, $\beta_p)$ denotes the vector of regression
coefficients. Then we can define a linear model for the outcome
so that
\begin{equation}
    Y_i = \beta_{T} T_i + \beta_0 + X\beta + \epsilon_i
\end{equation}
where $\epsilon_i\sim \mathcal{N}(0, \sigma^2)$. Clearly, when
the underlying true outcome model is linear,
\begin{equation}
    \delta = \beta_{T}\quad\text{and hence } 
    \hat{\delta} - \delta = \hat{\beta}_T - \beta_T.
\end{equation}

@@

\section{Robust Bayesian Analysis}
\subsection{Hierarchical model}

\begin{figure}
	\centering
	\begin{tikzpicture}[params/.style={circle, draw=black!60, very thick, minimum size=7mm},
	hyper/.style={circle, draw=black!60, fill=black!20, thick, minimum size=7mm},
	latent/.style={rectangle, draw=black!60, fill=black!10, dashed, minimum size=7mm},
	data/.style={rectangle, draw=black!60, thick, minimum size=8mm}]
	\node[params] (1) at (0,0) {$\pi$};
	\node[data] (2) at (3,0) {$X$};
	\node[data] (3) at (6,0) {$T$};
	\node[params] (4) at (1.5,1.5) {$\gamma$};
	\node[latent] (5) at (4.5,1.5) {$T^*$};
	\node[params] (6) at (1.5,-1.5) {$\beta$};
	\node[data] (7) at (3.5,-1.5) {$Y$};
	\node[hyper] (8) at (-1.5,-3) {$\sigma^2$};
	\node[params] (9) at (3,-3) {$\beta_0$};
	\node[params] (10) at (5.5,-1.5) {$\beta_T$};
	\node[hyper] (11) at (-1.5,-1) {$s$};
	\node[hyper] (12) at (-1.5,1) {$\alpha$};
	\draw[black, dashed] (-0.75,-3.75) rectangle (7,2.5);

	\path (1) edge[->]  (6);
	\path (1) edge[->]  (4);
	\path (8) edge[->]  (6);
	\path (6) edge[->]  (7);
	\path (2) edge[->]  (7);
	\path (2) edge[->]  (5);
	\path (5) edge[<-] (4);
	\path (5) edge[->] (3);
	\path (3) edge[->]  (7);
	\path (9) edge[->]  (7);
	\path (10) edge[->]  (7);
	\path (2) edge[dashed][->] (3);
	\path (8) edge[bend right = 60][->]  (10);
	\path (8) edge[->]  (9);
	\path (11) edge[->]  (1);
	\path (12) edge[->]  (1);

	\end{tikzpicture}
	\caption{Probabilistic graphical representation for causal inference with Bayesian hierarchical model.}
	\label{fig:regress}
\end{figure}

We perform a robust Bayesian analysis to obtain robust estimates. To do so, we propose the
following spike and slab priors \cite{ishwaran2005} to specify $\beta$ and $\gamma$.
For $1\le j\le p$,
\begin{align}
    \beta_j \mid \pi_{\beta}, \sigma^2 &\sim 
    \pi_{\beta}\normal\left(0, \tau_1^2\sigma^2\right) 
    + (1-\pi_{\beta}) \normal\left(0, \tau_0^2\sigma^2\right)\\
    \gamma_j \mid \pi_{\gamma} &\sim 
    \pi_{\gamma}\normal\left(0, \tau_1^2\right) 
    + (1-\pi_{\gamma}) \normal\left(0, \tau_0^2\right)\\
    \alpha\mid \sigma^2 &\sim \normal(0, \sigma^2)\\
    \sigma^2&\sim \text{InvGamma}(a, b)\\
    \pi_{\beta}, \pi_{\gamma}&\sim\text{Beta}\left(sq, s(1-q)\right).
\end{align}
We fix sufficiently small $\tau^2_0$
$(1\gg\tau_0^2>0)$ so that  $(\beta_j, \gamma_j) = (0,0)$ has its probability mass 
concentrated around zero. Therefore, this represents the spike component of our prior specification. 
To construct the slab component, we consider $\tau_1^2$ to be large so that 
$\tau_1^2\gg\tau_0^2$. This allows the prior for $(\beta_j, \gamma_j)\not=(0,0)$ to be flat.
